Bayesian exponential random graph modeling of KWL Burial network

### 2020-10-08 analysis/paper/paper.Rmd
edited method section and added references

<!--burial study- network analysis-->
Burials provide valuable information about past human behavior that can reflect not only the personal life and status of the deceased, but also the expressions of their descendants through grave goods and arrangements [@Pearson1999]. Mortuary practices such as burial forms, grave goods, and related ritual practices could represent the social relations between individuals and allow archaeologists to infer social structures. This enables the study of social differences based on burials to answer questions related to social inequality and complexity [@Ehrhardt2009; @Gamble2002; @Janes2013; @Johansen2015]. For examples, the use of burial goods can reflect social practices in a broader of cultural contexts to present personal wealth, social roles, or a traditional custom. Burials with grave goods in common often indicates underlying social relation connected through trade, exchange, and gifting [@Coward2013, pp. 252]. Identifying the structure of relationships among burials allows a further inferences on a structure of a social organization. One way to study relationships between individuals is social network analysis (SNA). A social network contains a set of actors and a set of relationships connecting these actors, which could be groups, organizations or persons. SNA can quantify, visualize, and statistically test hypotheses of the relationships between actors to examine the structure of network and major factors that affect the structure of the network [@Freeman2004; @Salvini2010]. To be specific, actors are "nodes" in a wider network, in which nodes are connected by multiple relationships, or "edges", to other nodes [@Coward2013; @Mizoguchi2013; @Potgieter2008]. Relationships can be viewed as flows of resources that reflect relations of control, dependence, and cooperation, reflecting social relation in a given society.

<!--archaeology case-->
The archaeological sites in northeastern Taiwan show some evidence of accumulation of imported prestige goods in burials around the same time as European contact in the 17th century [@Chen2007], which might indicate the pursuit of prestige or wealth in the society. In addition to prestige, European colonial powers could have provided military support to indigenous allies. For example, @Voss2005's study of a colonial outpost in California shows that colonial residents have different identities and higher social status than other local indigenous people. A general model to summarize these case studies might be that the support of a colonial power combined with high local values for imported goods, might lead to competition among individuals. In this study, we will use this model of competition to explain the changes in social inequality at Kiwulan, an Iron Age site in northesatern Taiwan, which covers the presence of the Europeans, including the Spanish and the Dutch, in the 17th century and the Chinese in the 19th century.

@@

<!--current approaches to network analysis-->
Exponential Random Graph Models (ERGMs) are an important family of statistical model for expressing properties of network structures with a focus directly on possible ties between nodes by viewing them as random variables [@Robins2007]. In traditional statistical model for network, each predictor variable for forming a tie between two nodes is measure separately by the presence of a tie without considering interactions among ties [@Morris2008]. By contrast, statistical models assumes social networks as dependent variables and allows a direct modeling for the network dependencies [@Snijders2011]. In a network generated by ERGMs, the formation of a tie is influenced by the presence or absence of other ties or attributes of nodes/ties. One can model network dependencies according to a common experience or research questions to represent relationships between actors through network variables. For example, reciprocity is a variable describing two actors' relationship, and transitivity describes two actors' relationship via a shared third actor. Those predictor variables, also called network statistics, are direct functions of the ties themselves where the probability of occurrence of ties can be specified and hypothesized [@Morris2008; @Robins2007]. The advantage of networks specification under assumptions and contexts is useful to explain archaeological network in a given time. For instance, @Amati2019 demonstrated the use of ERGMs to reconstruct three different network for 15 sites in the Caribbean between AD 100 to 400 to explore the interaction mechanisms. Those models test the hypotheses for the relations between sites in terms of proximity, inter-cultural, and exchange network to explain the presence of hub sites. They found that the observed presence of hubs can be efficiently explained by multiple interdependent mechanisms instead of the hypothesis of proximity or cultural homophily exclusively. However, they also point out some limitations of application, such as static outcome and sensitivity to missing data. In addition, the complex computational processes, such as intractable normalizing constant and degeneracy due to extreme graphs, limits ERGMs applications. A Bayesian framework allows for parameter inference using Markov chain Monte Carlo (MCMC) algorithm which avoid the need for computationally intensive calculations of the normalizing constants [@Caimo2011]. MCMC algorithms can produce approximate maximum likelihood estimators and graph statistics for further specification of transitivity and heterogeneity, this enables ERGMs to deal with complicated dependence patterns with a ease [@Snijders2006].

>>>>

<!--burial study- network analysis-->
Burials provide valuable information about past human behavior that can reflect not only the personal life and status of the deceased, but also the expressions of their descendants through grave goods and arrangements [@Pearson1999]. Mortuary practices such as burial forms, grave goods, and related ritual practices could represent the social relations between individuals and allow archaeologists to infer social structures. This enables the study of social differences based on burials to answer questions related to social inequality and complexity [@Ehrhardt2009; @Gamble2002; @Janes2013; @Johansen2015]. For examples, the use of burial goods can reflect social practices in a broader of cultural contexts to present personal wealth, social roles, or a traditional custom. Burials with grave goods in common often indicates underlying social relation connected through trade, exchange, and gifting [@Coward2013, pp. 252]. Identifying the structure of relationships among burials allows a further inferences on a structure of a social organization. A common way to study connections between individuals is social network analysis (SNA), which has been applied increasingly to archaeological studies [@Mills2017]. A network contains a set of actors and a set of relationships connecting these actors, which could be persons, groups, or organizations. In other words, actors are "nodes", in which nodes are connected by multiple relationships, or "edges", to other nodes. The basic unit for SNA analysis is dyad relationship, which is defined as a connection between two distinct actors [@Lee2016]. Relationships can be viewed as flows of resources that reflect relations of control, dependence, and cooperation, reflecting social relation in a given society [@Marin2011]. Using SNA, we can quantify, visualize, and statistically test hypotheses of the relationships between actors to examine the structure of network and major factors that affect the structure of the network [@Freeman2004; @Salvini2010]. The application in archaeology is widely used for [@Coward2013; @Mizoguchi2013; @Potgieter2008]

<!--archaeology case-->
The archaeological sites in northeastern Taiwan show evidence of use of imported prestige goods in burials and accumulation of them in residential area around the same time as European contact in the 17th century [@Chen2007], which might indicate the pursuit of prestige or wealth in the society. Those prestige goods include imported ornaments and porcelains that entered northeastern Taiwan through regional exchange networks involved the Europeans, Han Chinese traders, and Indigenous groups. In addition to prestige, European colonial powers could have provided military support to indigenous allies. For example, @Voss2005's study of a colonial outpost in California shows that colonial residents have different identities and higher social status than other local indigenous people. A general model to summarize these case studies might be that the support of a colonial power combined with high local values for imported goods, might lead to competition among individuals. In this study, we will use this model of competition to explain the changes in social inequality at Kiwulan, an Iron Age site in northesatern Taiwan, which covers the presence of the Europeans, including the Spanish and the Dutch, in the 17th century and the Chinese in the 19th century.

@@

<!--current approaches to network analysis-->
Exponential Random Graph Models (ERGMs) are an important family of statistical model for expressing properties of network structures with a focus directly on possible ties between nodes by viewing them as random variables [@Robins2007]. In traditional statistical model for network, each predictor variable for forming a tie between two nodes is measured separately by the presence of a tie without consideration of interactions among ties [@Morris2008]. By contrast, statistical models assumes social networks as dependent variables and allows a direct modeling for the network dependencies [@Snijders2011]. In a network generated by ERGMs, the formation of a tie is influenced by the presence or absence of other ties or the attributes of nodes/ties, called covariates. One can model network variables, such as dependencies, nodal attributes, and dyadic attributes, to generate edges between nodes based on a common experience for observed relationships. For example, common network dependencies include reciprocity that describes two actors' relationship, and transitivity that describes two actors' relationship via a shared third actor. Those variables, also called network statistics, are direct functions of the ties themselves where the probability of occurrence of ties can be specified and hypothesized [@Morris2008; @Robins2007]. The advantage of networks specification under assumptions and contexts is useful to explain archaeological network in a given time. For instance, @Amati2019 demonstrated the use of ERGMs to reconstruct three different network for 15 sites in the Caribbean between AD 100 to 400 to explore the interaction mechanisms. Those models test the hypotheses for the relations between sites in terms of proximity, inter-cultural, and exchange network to explain the presence of hub sites. They found that the observed presence of hubs can be efficiently explained by multiple interdependent mechanisms instead of the hypothesis of proximity or cultural homophily exclusively. However, they also point out some limitations of application, such as static outcome and sensitivity to missing data. In addition, the complex computational processes, such as intractable normalizing constant and degeneracy due to extreme graphs, limits ERGMs applications. A Bayesian framework allows for parameter inference using Markov chain Monte Carlo (MCMC) algorithm which avoid the need for computationally intensive calculations of the normalizing constants [@Caimo2011]. MCMC algorithms can produce approximate maximum likelihood estimators and graph statistics for further specification of transitivity and heterogeneity, this enables ERGMs to deal with complicated dependence patterns with a ease [@Snijders2006].
